---
title: "Investing using the CAPE Ratio: An Analysis Using R"
author: "Karna Patel"
date: "15/10/2020"
output:
  html_document: default
  pdf_document: default
---


**The purpose of this analysis is to assess the accuracy of the CAPE ratio in predicting long term stock market returns.This analysis will include a summary and introduction to this topic, commentary on various relevant academic literatures, analysis of the CAPE ratio and prior market activity, and the synthesis and evaluation of a simple investment strategy using this metric.**

**The cyclical adjusted price-to-earnings (CAPE or Shiller’s P/E) ratio is a variation of the price-to-earnings (P/E) ratio. This concept was first introduced by Robert Shiller and John Campbell, who proposed calculating the P/E ratio with an average of real earnings from the previous ten years.**


**Before creating our investment model, preliminary data analysis was conducted on S&P market data, CAPE ratios, and other economic indicators. Our team wanted to find a clean dataset containing all the needed information for our analysis within it. Shiller’s public dataset on real market returns, real interest rates, inflation levels and CAPE ratio values was used.  This dataset provides monthly data on S&P index prices, S&P index dividends, S&P index earnings, inflation, long-term interest rates, market returns, and CAPE ratios for the period from January 1871 to August 2020, which provided a strong basis for our analysis. Prior to creating our model, preliminary cleaning for unavailable data and data analysis was conducted on market data, including conducting regression analyses, computing long-term figures, and conducting other simple data calculations.**

## Loading & Cleaning Data


```{r setup}
knitr::opts_chunk$set(echo = FALSE)

#loading all relevant library files
library(tidyverse)  
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(stringr)
library(TTR)
library(zoo)
```



```{r one, echo=TRUE}
# reading csv file of preliminary data
dataset <- read_csv("projectdata.csv")


#  data cleaning on our CSV file.
dataset2 <- dataset %>% 
            filter(Year >= 1900 ) %>%
               # cutting data before 1900
            mutate(                     # changing months from numbers to charachters
                    Month = case_when(
                    Month == 1 ~ "January",
                    Month == 2 ~ "February",
                    Month == 3 ~ "March",
                    Month == 4 ~ "April",
                    Month == 5 ~ "May",
                    Month == 6 ~ "June",
                    Month == 7 ~ "July",
                    Month == 8 ~ "August",
                    Month == 9 ~ "September",
                    Month == 10 ~ "October",
                    Month == 11 ~ "November",
                    Month == 12 ~ "December"))

dataset2   # looking at data to see what it looks like

# Now, we will plot the CAPE ratio to see what it looks like over the selected 70 year period.

plot_1 <- dataset2 %>% ggplot(aes(x=Date))+
                      geom_line(aes (y=`CAPE (P/E10)`),
                                     color = "black", 
                                     size = 1)+
                      geom_line(aes (y=mean(`CAPE (P/E10)`)),
                                     color = "maroon", 
                                     size = 1)+          
                      labs(title = "Exhibit 1: Mapping the Historical Cape Ratio", 
                           subtitle = "S&P500's Historical CAPE Ratio alongside Historical Average", 
                           caption = "Data Source: Princeton University Press, Robert J. Shiller ",
                           y = "CAPE Ratio",
                           x = "Year") +
                      theme_bw(base_size = 10) +
                      geom_label(x=1999,
                                y=45, 
                                label ="2000 Dot-Com Bubble", 
                                size = 3) +
                      geom_label(x=1929,
                                y=35, 
                                label ="1929 Stock Market Bubble", 
                                size = 3)+
                      geom_label(x=2008,
                                y=10, 
                                label ="2008 U.S. Housing Crisis", 
                                size = 3)
plot_1



```

## Analyzing Data 

```{r two, echo=TRUE}

# We want to confirm the accuracy of the provided data by doing our own calculation on the CAPE ratio.

dataset3 <- dataset2 %>%  
            filter(Decade >= 1950 | Decade <= 2010) %>%
            mutate(Calculated_CAPE = `Real Prices` / `BRolling Average Real Earnings`) %>%
            group_by(Decade)
            # summarise(Median = median(Calculated_CAPE), `CAPE (P/E10)`,`Year`, `Decade`)
            
dataset3


#Creating graph to show Cape Ratio vs. Market Returns following 10 Years

plot_2 <-ggplot(data = dataset3, aes(x = `CAPE (P/E10)`, 
                            y = `10Y FRolling Return Percentage`, 
                            color = Decade)) +
              geom_point() +
              labs(title = "Exhibit 3: The CAPE Ratio is a Good Predictor of Long-Term Market Returns...", 
                   subtitle = "CAPE Ratios and Subsequent 10 Year Real Market Returns", 
                   caption = "Data Source: Princeton University Press, Robert J. Shiller", 
                   y = "Next 10Y Average Annual Return",
                   x = "CAPE Ratio") +
              theme_bw(base_size = 10) +
              scale_x_continuous() +
              scale_y_continuous() + 
              geom_smooth(method="lm",color = "darkred") +
              guides(colour = guide_colourbar(order = 1))

plot_2


#Creating graph to show Cape Ratio vs. Market Returns following 1 Years

plot_3 <-ggplot(data = dataset3, aes(x = `CAPE (P/E10)`, 
                            y = `1Y FRolling Return Percentage`, 
                            color = Decade)) +
              geom_point()+
              labs(title = "Exhibit 4: The CAPE Ratio is NOT a Good Predictor of Short-Term Market Returns...", 
                   subtitle = "CAPE Ratios and Subsequent 1 Year Real Market Returns", 
                   caption = "Data Source: Princeton University Press, Robert J. Shiller", 
                   y = "Next 1Y Annual Return",
                   x = "CAPE Ratio") +
              theme_bw(base_size = 10) +
              scale_x_continuous() +
              scale_y_continuous() + 
              geom_smooth(method="lm",color = "darkred")+
            scale_fill_gradientn(colours = terrain.colors(6))

plot_3





```




```{r three, echo=TRUE}



# Analyzing Median Cape Ratios to see Trends since 1950
dataset3


dataset4 <- dataset3 %>%
            filter(Decade <= 2010) %>%
            group_by(Decade) %>%
            # summarise(Max = max(`Rolling 10Y CAPE Average`), 
            #           Min = min(`Rolling 10Y CAPE Average`), 
            #           Median = median(`CAPE (P/E10)`)) %>%
            arrange(Year)


dataset4

#Moving Average Graphs

decade_graph <- ggplot(dataset4, aes(x = `Year`,y=`CAPE (P/E10)`))+
                      geom_line(size=1)+
                      geom_smooth(method="lm", 
                                  color = "maroon", 
                                  size = 1)+
                      geom_quantile()+
                      labs(title = "Exhibit 2:  CAPE median for Drifting Valuations over the past 70 years", 
                                 subtitle = "CAPE Ratio Moving Average with Quantiles", 
                                 caption = "Data Source: Princeton University Press, Robert J. Shiller",
                                 y = "CAPE Ratio",
                                 x = "Year") +
                      theme_bw(base_size = 10)
decade_graph


decade_graph2 <- ggplot(dataset4, aes(x = `Year`,y=`CAPE (P/E10)`))+
                      geom_line(size=1)+
                      geom_smooth(method="lm", 
                                  color = "maroon", 
                                  size = 1)+
                      facet_grid(.~Decade, scales = "free_x")+
                      labs(title = "Exhibit 5: How does the CAPE ratio median change from decade to decade?", 
                                 subtitle = "CAPE Ratio Moving Median by Decades", 
                                 caption = "Data Source: Princeton University Press, Robert J. Shiller",
                                 y = "CAPE Ratio",
                                 x = "Year") +
                      theme_bw(base_size = 10)+
                       scale_x_continuous(breaks = c())
decade_graph2



# median CAPE ratios graph
d4 <- dataset4 %>% filter(Decade <= 2000) %>%
            group_by(Decade) %>%
            summarise(Median = median(`CAPE (P/E10)`)) 
d4

g4 <- ggplot (d4,aes (x=`Decade`,y = `Median`)) +
                  geom_col()+
                  labs(title = "Exhibit 6: How have Median CAPE ratios changed?", 
                           subtitle = "Median 10Y CAPE", 
                           caption = "Data Source: Princeton University Press, Robert J. Shiller ",
                           y = "Median CAPE ",
                           x = "Decade")+ 
                  theme_bw()+
                  geom_label(inherit.aes = TRUE, label = d4$Median)

g4




```

```{r four, echo=TRUE}


#Returns graph

returns <- read_csv("returnsdata.csv")
returns

returns_plot <- returns %>% ggplot(aes(x=Date)) +
                      geom_line(aes(y=`CAPE Portfolio`, 
                                    color="CAPE Portfolio"))+
                      geom_line(aes(y=`Buy and Hold Portfolio`, 
                                    color="Buy and Hold Portfolio"))+
                      # facet_grid(~Year,scales = "free_x")+
                      labs(title = "Exhibit 7: How does the Strategy Perform?", 
                           subtitle = "Portfolio Value of $1M invested in CAPE Ratio Strategy vs. Buy & Hold Strategy", 
                           caption = "Data: Our Calculated Returns Analysis ",
                           y = "Portfolio Value",
                           x = "Year") +
                      theme_bw(base_size = 10)+
                      scale_y_continuous()+
                      scale_x_continuous()+
                      geom_label(x=2015,
                                y=1750000, 
                                label ="CAPE Portfolio: $1,770,296", 
                                size = 3)+
                      geom_label(x=2018,
                                y=1500000, 
                                label ="BH Portfolio: $1,823,294", 
                                size = 3)+ 
                        theme(legend.position = "none")

returns_plot
```  

**The model has given us a strong understanding of the implications of using the CAPE ratio as a metric for predicting future capital market returns. While it is a great theoretical financial metric, it is important to consider the context in which it is being used. It is not an all-encompassing metric that will always signal when a market is undervalued or overvalued. Similar to any other measure, it also has drawbacks and weaknesses that should be understood when using it in a practical sense and it is only as accurate as the numbers used when calculating it. A singular focus on using this metric can result in investors failing to accurately understand the market and missing out on opportunities to generate returns. As our strategy demonstrated, the CAPE ratio is still a phenomenal metric that investors should use and when adjusted for its shortcomings, it can be a great tool to analyze and understand the market.**

