---
title: "GDP components over time and among countries"
author: "Karna Patel"
date: "21/10/2020"
output: html_document
---
**How has GDP among different countries changed over time? I hope to answer this question through the following analysis done on R, including a breakdown on the trends of a couple specific countries. While data analysis is always important, I will try to understand what the identified trends are and what has caused them.**

**The main components of gross domestic product, GDP are personal consumption (C), business investment (I), government spending (G) and net exports (exports - imports).We will look at how GDP and its components have changed over time, and compare different countries and how much each component contributes to that country's GDP.**



```{r setup }
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	fig.height = 6.75,
	fig.width = 6.75,
	message = FALSE,
	warning = FALSE,
	size = "small",
	tidy = FALSE
)
   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)


library(tidyverse)  
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(stringr)
library(TTR)
library(zoo)
```




```{r read_GDP_data, eval=TRUE}

UN_GDP_data  <-  read_excel(("Download-GDPconstant-USD-countries.xls"), # Excel filename
                sheet="Download-GDPconstant-USD-countr", # Sheet name
                skip=2) # Number of rows to skip

```



```{r reshape_GDP_data}

UN_GDP_data
skim(UN_GDP_data)



tidy_GDP_data  <- UN_GDP_data %>%
                  pivot_longer(cols=4:51,names_to="Year",values_to = "Expenditure") %>%
                  mutate (Expenditure = Expenditure / 1000000000 ) %>%
                  mutate(
                        IndicatorName = case_when(
                        IndicatorName == "Final consumption expenditure" ~ "FCI",
                        IndicatorName == "Household consumption expenditure (including Non-profit institutions serving households)" ~ "Household Expenditure",
                        IndicatorName == "General government final consumption expenditure" ~ "Government Expenditure",
                        IndicatorName == "Gross capital formation" ~ "Gross capital formation",
                        IndicatorName == "Gross fixed capital formation (including Acquisitions less disposals of valuables)" ~ "GFCF",
                        IndicatorName == "Exports of goods and services" ~ "Exports",
                        IndicatorName == "Imports of goods and services" ~ "Imports",
                        IndicatorName == "Gross Domestic Product (GDP)" ~ "GDP",
                        IndicatorName == "Agriculture, hunting, forestry, fishing (ISIC A-B)" ~ "AHFF",
                        IndicatorName == "Mining, Manufacturing, Utilities (ISIC C-E)" ~ "MMU",
                        IndicatorName == "Manufacturing (ISIC D)" ~ "M",
                        IndicatorName == "Construction (ISIC F)" ~ "C",
                        IndicatorName == "Wholesale, retail trade, restaurants and hotels (ISIC G-H)" ~ "WRRH",
                        IndicatorName == "Transport, storage and communication (ISIC I)" ~ "TSC",
                        IndicatorName == "Other Activities (ISIC J-P)" ~ "OA",
                        IndicatorName == "Total Value Added" ~ "TVA"))
     

tidy_GDP_data


# Let us compare GDP components for these 3 countries
country_list <- c("United States","India", "Germany")
```



```{r gdp1, echo=FALSE, out.width="100%"}


country_list


#cleaning data
tidy_GDP_data2 <- tidy_GDP_data %>%
                    group_by(Year) %>%
                    select(Country, 
                           Year, 
                           IndicatorName, 
                           Expenditure) %>%
                    filter(Country == "Germany" | Country == "India" | Country == "United States") %>%
                    filter(IndicatorName %in% c("Exports",
                                                "Imports",
                                                "Gross capital formation",
                                                "Government Expenditure",
                                                "Household Expenditure")) 
tidy_GDP_data2

#plotting a graph
ggplot (tidy_GDP_data2, aes (x = Year, 
                             y = Expenditure, 
                             color = IndicatorName, 
                             group = IndicatorName)) + 
        geom_line()+
        labs ( y = "Billion US$", 
               title = "GDP components over time", 
               subtitle = "In Constant USD 2010", x ="", 
               legend = "Components of GDP",
               IndicatorName = "Components of GDP") +             
        facet_wrap(~ Country) +
        theme_bw(base_size = 8) +
        scale_x_discrete(breaks = c(1970,1980,1990,2000,2010,2020)) +
        guides(color=guide_legend("Components of GDP"))

```



```{r gdp3, echo=FALSE, out.width="100%"}
#Recalculating GDP

tidy_GDP_data

tidy_GDP_data3 <- tidy_GDP_data %>%
                    group_by(Year, Country) %>%
                    arrange(Year, Country) %>%
                    select(Country, Year, IndicatorName, Expenditure) %>%
                    filter(IndicatorName %in% c("Exports",
                                              "Imports",
                                              "Gross capital formation",
                                              "Government Expenditure",
                                              "Household Expenditure", 
                                              "GDP")) %>%
                    rowwise() %>%
                    mutate(CalculatedGDP = sum (IndicatorName == "Household Expenditure",
                        IndicatorName == "Government Expenditure",
                        IndicatorName =="Gross capital formation",
                        IndicatorName == "Exports",
                        IndicatorName == "Imports"))



tidy_GDP_data3 

```

The last chart breaks down the GDP of Germany, India, and the United States. It maps how much Government Expenditure, Gross Capital Formation, Net Exports, and Household Expenditure have contributed to total GDP over the years.

**For Germany, their Household expenditure has dipped in its contribution to aggregate GDP as has Gross Capital Formation. Their Government Expenditure has stayed relatively the same in terms of contribution. However, Net Exports has increased signficantly. The increase can primarily be seen beginninng in the 1990s into the 2000s. I believe a plausible explanation for this is the end of the Soviet Union and the reunification between East and West Germany. As the economy moved away from communism and opened up economic borders, the economy opened up and the country’s net exports increased significantly.**

**For India, Household Expenditure has decreased its contribution into GDP significantly, while Government Expenditure and Net Exports (for the most part) have stayed the same. Over the past 20 years, Gross Capital Formation has heavily increased its contribution to aggregate GDP.I think a plausible explanation for this trend can be seen through the economic policies of the government. In the 1970s and 1980s, the government of India’s economic policies were not favoured towards trading or pursuing foreign relations (influenced by close relations with the Soviet Union). Since the 1990s into the 2000s, the government has ditched those past policies in favour of more open borders, trade agreements, and globalisation. This has resulted in a boom of foreign investment into the region (“oursourcing”) as well as better domestic business growth, which is seen through gross capital formation’s increased contribution. For household expenditure, their overall figures might be the same, but the contribution has been coming down due to increases in the total aggregate gdp from other sources. One could also argue that their household contribution to GDP is coming down to what other countries around the world see with their GDP.**

**For the United States, Net Exports and Government Expenditure have been slowly decreasing, while Household Expenditure and Gross Capital Formation have slighty increased. I think these trends make sense given their economy’s switch away from manufacturing to a more service-oriented, which means their net exports have decreased. Further, their economy has been strong (on average) for the past 50 years and their households (again on average) some of the wealthiest in the world, which is why these two factors have continued to increase their contribution towards GDP.**

```{r gdp2, echo=FALSE, out.width="100%"}

#plotting a graph for Canada, Mexico, and United States

country_list2 <-c("United States","India", "Germany")


tidy_GDP_data4 <- tidy_GDP_data %>%
                    group_by(Year) %>%
                    select(Country, Year, IndicatorName, Expenditure) %>%
                    filter(Country == "Canada" | Country == "Mexico" | Country == "United States") %>%
                    filter(IndicatorName %in% c("Exports", "Imports", "Gross capital formation", "Government Expenditure", "Household Expenditure" ))
                    
                    

tidy_GDP_data4

ggplot (tidy_GDP_data4, aes (x = Year, y = Expenditure, color = IndicatorName, group = IndicatorName)) + 
        geom_line()+
        labs ( y = "Billion US$", title = "GDP components over time", subtitle = "In Constant USD 2010" ) +             facet_wrap(~ Country) +
        theme_bw(base_size =8) +
        scale_x_discrete(breaks = c(1970,1980,1990,2000,2010,2020))+
        guides(color=guide_legend("Components of GDP"))
      
```
**Why do you think the trends for the US, Canada, and Mexico are what they are? Feel free to leave a comment or email me, I would love to hear what your thoughts are!**
